---
title: "The Data Incubator - Challenge"
author: "Jake Johnson"
date: "10/29/2018"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = 'figures/README-'
  )
load_tidy()
library(hrbrthemes)
```

## Proposal

One prediction of climate change is that the intensity and rainfall rates of cyclones will increase[^1].  Also expected to increase are the damages and costs of such storms.  In the U.S., hurricanes Florence, Harvey, and Michael serve as the most recent examples.

The devastating effects of these events are demonstrated starkly via the use of satellite imagery both before and after they occur.  For example, this Washington Post [graphic](https://www.washingtonpost.com/graphics/2018/world/amp-stories/before-and-after-images-of-typhoon-yutu-destruction/) shows the destruction brought by Super Typhoon Yutu.

Recently, efforts have been made to use such imagery as means for automated damage detection in storm-effected areas[^2][^3].  I propose using and buidling on these existing methodologies to assess storm-damaged areas and linking this information to estimates of localized monetary damages.  This would provide a novel approach for predicting the costs of future storms.

## Data Sources

- [National Hurricane Center](https://www.nhc.noaa.gov/data/hurdat/)

- [Billion-Dollar Disasters](https://www.ncdc.noaa.gov/billions/time-series)

- [National Geodetic Survey](https://www.ngs.noaa.gov/)
  + [Emergency Response Imagery - Before](https://geodesy.noaa.gov/storm_archive/coastal/viewer/index.html)
  + [Emergency Response Imagery - After](https://storms.ngs.noaa.gov/)
  
- [FEMA Spending](https://www.usaspending.gov/#/)

- [FEMA National Disaster Assessments](https://data.femadata.com/NationalDisasters/)

## Plots

```{r plot-one, echo = F, message = F, warning = F}
dat <- read_csv(
  'raw/hurricanes/hurdat2-atl-1851-2017-050118.txt',
  col_names = F
  )

dat %<>% filter(is.na(X4))
dat %<>%
  mutate(
    year = str_sub(X1, 5, 8)
  )

ggplot(dat) +
  geom_bar(aes(x = year)) +
  theme_ipsum() +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(
    title = "Atlantic Hurricanes 1851 to 2017",
    x = "Year", y = "Count",
    caption = "Source: NOAA"
  )
```

```{r plot-two, echo = F, message = F, warning = F}
dat <- read_csv(
  "raw/hurricanes/noaa-billions-time-series.csv",
  skip = 1
  )
dat %<>% select(Year, starts_with("Tropical Cyclone"))
names(dat) %<>% tolower()
names(dat) %<>% str_replace_all(" ", "_")

ggplot(dat, aes(x = year)) +
  geom_point(
    aes(y = tropical_cyclone_cost)
  ) +
  geom_errorbar(
    aes(ymin = tropical_cyclone_lower_95, ymax = tropical_cyclone_upper_95)
  ) +
  theme_ipsum() +
  labs(
    title = "Billion-dollar hurricanes by year",
    x = "Year", y = "Billions USD",
    caption = "95 error bars shown.  Source: NOAA"
  )
```

```{r plot-three, echo = F, message = F, warning = F}
dat <- read_xlsx(
  "raw/hurricanes/Census Tract Damage Assessment.xlsx"
  )
dat %<>% gather(type, value, -Key)

ggplot(filter(dat, type != "Total")) +
  geom_bar(
    aes(x = Key, y = value, fill = type), stat = "identity"
  ) +
  scale_fill_discrete(
    labels = c(
      "Affected",
      "Destroyed",
      "Major damage",
      "Minimal damage"
    )
  ) +
  theme_ipsum() +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(
    title = "Tract-level damage assessment from Hurricane Harvey",
    x = "Census tracts", y = "Counts",
    caption = "Source: http://www.spgroupusa.com/"
  )
```


[^1]:
https://www.gfdl.noaa.gov/global-warming-and-hurricanes/

[^2]:
Barnes, C. F., Fritz, H., & Yoo, J. (2007). Hurricane Disaster Assessments With Image-Driven Data Mining in High-Resolution Satellite Imagery. IEEE Transactions on Geoscience and Remote Sensing, 45(6), 1631–1640. http://doi.org/10.1109/TGRS.2007.890808

[^3]:
Cao, Q. D., & Choe, Y. (2018, July 4). Deep Learning Based Damage Detection on Post- Hurricane Satellite Imagery. arXiv.